---
title: "Final Results"
author: "Frank"
format: pdf 
---

```{r}
#| message: false
library(sf)
library(tidyverse)

# Load shapefiles merged with lead data
load('data/bllWorld.RData')
load('data/bllGBD.RData')

# Set ggplot theme
theme_set(theme_bw())
```

# Fraction >5 and >10

```{r}
# Fraction with BLL > 5 micrograms / deciliter
bllWorld |> 
  ggplot() + 
  geom_sf(aes(fill = 100 * frac5plus)) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt", name = '') +
  labs(title = 'Estimated % of population aged 0-19 with BLL > 5μg/dl')
```

```{r}
# Fraction with BLL > 10 micrograms / deciliter
bllWorld |> 
  ggplot() + 
  geom_sf(aes(fill = 100 * frac10plus)) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt", name = '') +
  labs(title = 'Estimated % of population aged 0-19 with BLL > 10μg/dl')
```

# Average IQ Loss per individual aged 0-19: Beta approximation

## IQ Points
```{r}
# Average IQ lost per individual aged 0-19: beta approximation
bllWorld |> 
  ggplot() + 
  geom_sf(aes(fill = beta_IQ_integral)) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt")
```

## Dollar Value
This version uses returns to education. Notice that some of the countries with the worst lead problem are missing returns data, hence we don't have a dollar value here:

```{r}
# $ value of IQ lost per individual aged 0-19: beta approximation
bllWorld |> 
  ggplot() + 
  geom_sf(aes(fill = beta_returns_no_lead)) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt")
```
This version uses a figure of \$1000 per year per IQ point per person. I calculated this by discounting the lifetime figure from our other paper of around \$20,000 over a 40 year working lifetime with a discount rate between 3 and 4\%. This addresses the missing data problem.

```{r}
# Beta 
bllWorld |> 
  ggplot() + 
  geom_sf(aes(fill = beta_IQ_integral * 1000)) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt")
```

# Average IQ loss per individual aged 0-19: Lognormal approximation

## IQ Points
```{r}
bllWorld |> 
  ggplot() + 
  geom_sf(aes(fill = lnorm_IQ_integral)) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt")
```

## Dollar value
This version uses returns to education. Notice that some of the countries with the worst lead problem are missing returns data, hence we don't have a dollar value here:

```{r}
# $ value of IQ lost per individual aged 0-19: lognormal approximation
bllWorld |> 
  ggplot() + 
  geom_sf(aes(fill = lnorm_returns_no_lead)) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt")
```

This version uses a figure of \$1000 per year per IQ point per person. I calculated this by discounting the lifetime figure from our other paper of around \$20,000 over a 40 year working lifetime with a discount rate between 3 and 4\%.

```{r}
# Lognormal 
bllWorld |> 
  ggplot() + 
  geom_sf(aes(fill = lnorm_IQ_integral * 1000)) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt")
```


# Average IQ loss per individual aged 0-19: lower bound approach

Assign BLL of 10 to everyone above 10, 5 to everyone between 5 and 10, and zero to everyone else. 

## IQ Points

```{r}
# Average IQ lost per individual aged 0-19: lower bound 
bllWorld |> 
  ggplot() + 
  geom_sf(aes(fill = LB_IQ_integral)) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt")
```

## Dollar value
This version uses returns to education. Notice that some of the countries with the worst lead problem are missing returns data, hence we don't have a dollar value here:
```{r}
# $ value of IQ lost per individual aged 0-19: lower bound 
bllWorld |> 
  ggplot() + 
  geom_sf(aes(fill = LB_returns_no_lead)) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt")
```

This version uses a figure of \$1000 per year per IQ point per person. I calculated this by discounting the lifetime figure from our other paper of around \$20,000 over a 40 year working lifetime with a discount rate between 3 and 4\%. This results in much less missing data:

```{r}
# Lower bound
bllWorld |> 
  ggplot() + 
  geom_sf(aes(fill = LB_IQ_integral * 1000)) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt")
```

# How different are the Beta, Lognormal, and Lower Bound approaches?

```{r}
# | message: false
# | warning: false
library(GGally)
bllGBD |> 
  select(LB_IQ_integral, beta_IQ_integral, lnorm_IQ_integral) |> 
  ggpairs()
bllGBD |> 
  select(LB_IQ_integral, beta_IQ_integral, lnorm_IQ_integral) |> 
  cor()
```

# Various Ways of Calculating the Value of Removing Lead

There's a small issue we need to think about when comparing results across fixed and variable IQ costs: we have fewer countries for the latter exercise, so the figures will necessarily be smaller. Not sure of the best way to handle this. 

We also need to make sure we're clear on the units here. It's per year but for the entire cohort of 0-19 year olds. This is a little bit tricky to think about. I think it might be clearer to express this in per person terms. This also resolves the issue regarding missing data: we simply need to normalize by total population among the group of countries for which we have data.

```{r}
#-------------------------------------------------------------------------------
# IQ points lost as a function of BLL (Lanphear, 2005)
#-------------------------------------------------------------------------------
iq_loss <- function(bll) {
  # In our other paper we multiplied this by 20568 to convert to a dollar cost.
  # Here we leave the result is in units of IQ points.
  pmin(bll, 10) * 0.513 + (bll >= 10) * pmin(bll - 10, 20 - 10) * 0.19 +
             (bll >= 20) * (bll - 20) * 0.11
}
```


## Estimate the $ benefit from eliminating lead: Version 1# 
 In our IL paper we monetized the loss of one IQ point for a 3-year old as as 20,568 in 2019 dollars, following Klemick, Mason, and Sullivan (2020). Their calculation was the present value of total lifetime earnings loss in the US. This value may not apply outside the US, but could serve as a simple benchmark. 
 
If we assume a discount rate between 3 and 4% per year, along with a working lifetime between 40 and 45 years, we need to deflate the total figure by a factor of between 20 and 25. This comes to roughly $1000 per year. We want a flow so that the figure is comparable to the one we compute below.

```{r}
bllGBD |> 
  summarize(lower_bound = sum(1000 * popn0019 * LB_IQ_integral),
            beta_approx = sum(1000 * popn0019 * beta_IQ_integral),
            lnorm_approx = sum(1000 * popn0019 * lnorm_IQ_integral))
```

The beta and lognormal both give a figure of around 5-6 trillion. The lower bound approach is about half as big: 2.5 trillion. No matter how you look at it, these are big numbers. Take the lower bound and divide it by 2 on the premise that part of the estimated relationship between lead and IQ is endogenous family background: you still get more than 1 trillion per year. Compare to world GDP of approximately 96 trillion in 2021.


Try re-expressing this in per-person terms: *divide* by total population between 0 and 19 at the end. This gives a weighted average based on popn.

```{r}
bllGBD |> 
  summarize(lower_bound = sum(1000 * popn0019 * LB_IQ_integral) / sum(popn0019),
            beta_approx = sum(1000 * popn0019 * beta_IQ_integral) / sum(popn0019),
            lnorm_approx = sum(1000 * popn0019 * lnorm_IQ_integral) / sum(popn0019))
```

These figures are much easier to think about: 974 for the lower bound, 2221 for the beta and 1947 for the lognormal. Compare to world GDP / capita of around 12,234. 

## Estimate the $ benefit from eliminating lead: Version 2

This approach combines estimated returns to education across a large number of countries with GDP per capita at PPP. The idea is that GDP / capita at PPP is an approximation of income / capita. Multiplying this value by estimated returns to one additional year of education gives a result in \$/year per additional year of education. To convert this to \$/year per IQ point, we need the relationship between IQ points and years of education. 
```{r}
bllGBD |> 
  filter(!is.na(beta_returns_no_lead)) |> 
  summarize(lower_bound = sum(LB_returns_no_lead * popn0019), 
            beta_approx = sum(beta_returns_no_lead * popn0019), 
            lnorm_approx = sum(lnorm_returns_no_lead * popn0019))
```

These figures are broadly in line with what we found above: now we have 1.6 trillion for the lower bound, 8 for the beta approximation and 4.5 for the lognormal approximation. Compare to World GDP in 2021 of approximately 96 trillion.

Now put them in per person terms, dividing by total population 0 to 19
```{r}
bllGBD |> 
  filter(!is.na(beta_returns_no_lead)) |> 
  summarize(lower_bound = sum(LB_returns_no_lead * popn0019) / sum(popn0019), 
            beta_approx = sum(beta_returns_no_lead * popn0019) / sum(popn0019), 
            lnorm_approx = sum(lnorm_returns_no_lead * popn0019) / sum(popn0019))
```
664 for lower bound, 3330 for beta approx and 1868 for lognormal. Compare to world GDP per capita of around 12,000.

## Sanity check: histogram of the implied dollars per IQ point per year
```{r}
bllGBD |> 
  ggplot(aes(x = dollar_per_IQ)) +
  geom_histogram() +
  scale_x_log10() +
  xlab('Implied Dollars per IQ Point Per Year Across Countries')

bllGBD |> 
  filter(!is.na(dollar_per_IQ)) |> 
  summarize(median(dollar_per_IQ), mean(dollar_per_IQ))
```

Notice that the median is \$787 and the mean is \$2000. These bookend the \$1000 figure that we calculated above by discounting the \$20,000 lifetime figure.

## Alternative Counterfactual
The counterfactuals given above assume that we take everyone down to a BLL of zero. What if we instead brought everyone down to 5? There are many ways to calculate this. A simple lower bound assigns a BLL of 10 to everyone with a BLL > 10 and lowers this to 5.

```{r}
iq_gain <- iq_loss(10) - iq_loss(5)
bllGBD |> 
  mutate(total_IQ_gain = iq_gain * total10plus) |> 
  summarize(total = sum(1000 * total_IQ_gain), 
            per_person = sum(1000 * total_IQ_gain) / sum(popn0019))

bllGBD |> 
  mutate(total_IQ_gain = iq_gain * total10plus) |> 
  filter(!is.na(dollar_per_IQ)) |> 
  summarize(total = sum(dollar_per_IQ * total_IQ_gain), 
            per_person = sum(dollar_per_IQ * total_IQ_gain) / sum(popn0019))
```

These are comparatively smaller numbers 173 and 59 per person or 451 billion and 141 billion total. But this is *really* a lower bound because it assumes no one is above 10.



